---
title: "PM566-Assignment2"
author: "YangXu"
format:
  html:
    embed-resources: true
---
# Read in the data
```{r, message=FALSE}
library(tidytext)
library(tidyverse)
library(data.table)
library(dplyr)
library(dtplyr)
library(leaflet)
library(ggplot2)
library(data.table)
library(cowplot)
library(viridis)
chs_ind <- data.table::fread("chs_individual.csv")
chs_reg <- data.table::fread("chs_regional.csv")
```

# Data Wrangling
## Merge the data
```{r}
chs <- merge(chs_ind, chs_reg, by = "townname", all.x = TRUE)
dim(chs_ind)
dim(chs_reg)
dim(chs)
```
From the nrows and ncols, the merge is successful.

## 1. In the case of missing values, impute data using the average amongst individuals with the same values for the “male” and “hispanic” variables. For categorical variables, take the mode.
```{r}
summary(chs)
# There is no N/A in male or hispanic. So let's divide chs into four datasets by combination of male and hispanic.
chs_00 <- chs[male == "0" & hispanic == "0",]
chs_01 <- chs[male == "0" & hispanic == "1",]
chs_10 <- chs[male == "1" & hispanic == "0",]
chs_11 <- chs[male == "1" & hispanic == "1",]
# Then impute the missing value in each four categories.
# Now we only consider bmi, fev, smoke and gasstove
# bmi
chs_00$bmi[is.na(chs_00$bmi)] <- mean(chs_00$bmi[!is.na(chs_00$bmi)])
chs_01$bmi[is.na(chs_01$bmi)] <- mean(chs_01$bmi[!is.na(chs_01$bmi)])
chs_10$bmi[is.na(chs_10$bmi)] <- mean(chs_10$bmi[!is.na(chs_10$bmi)])
chs_11$bmi[is.na(chs_11$bmi)] <- mean(chs_11$bmi[!is.na(chs_11$bmi)])
# fev
chs_00$fev[is.na(chs_00$fev)] <- mean(chs_00$fev[!is.na(chs_00$fev)])
chs_01$fev[is.na(chs_01$fev)] <- mean(chs_01$fev[!is.na(chs_01$fev)])
chs_10$fev[is.na(chs_10$fev)] <- mean(chs_10$fev[!is.na(chs_10$fev)])
chs_11$fev[is.na(chs_11$fev)] <- mean(chs_11$fev[!is.na(chs_11$fev)])
# For smoke and gastove, we need to get the mode
# smoke
table(chs_00$smoke) # 0
table(chs_01$smoke) # 0
table(chs_10$smoke) # 0
table(chs_11$smoke) # 0

chs_00$smoke[is.na(chs_00$smoke)] <- 0
chs_01$smoke[is.na(chs_01$smoke)] <- 0
chs_10$smoke[is.na(chs_10$smoke)] <- 0
chs_11$smoke[is.na(chs_11$smoke)] <- 0

# gasstove
table(chs_00$gasstove)  # 1
table(chs_01$gasstove)  # 1
table(chs_10$gasstove)  # 1
table(chs_11$gasstove)  # 1

chs_00$gasstove[is.na(chs_00$gasstove)] <- 1
chs_01$gasstove[is.na(chs_01$gasstove)] <- 1
chs_10$gasstove[is.na(chs_10$gasstove)] <- 1
chs_11$gasstove[is.na(chs_11$gasstove)] <- 1

# bind them into chs
chs <- rbind(chs_00,chs_01,chs_10,chs_11)
summary(chs)

```
There is no more N/As in these variables
## 2. Create a new categorical variable named “obesity_level” using the BMI measurement (underweight BMI<14; normal BMI 14-22; overweight BMI 22-24; obese BMI>24).
```{r}
chs$obesity_level <- as.factor (ifelse(chs$bmi < 14, 'underweight',
                                ifelse(chs$bmi < 22, 'normal', 
                                ifelse(chs$bmi < 24, 'overweight','obese' ))))

# create a summary table that contains the minimum BMI, maximum BMI, and the total number of observations per category.
chs %>%
  group_by(obesity_level) %>%
  summarize(min_bmi = min(bmi),
            max_bmi = max(bmi),
            Count = n()) %>%
  knitr::kable()
```

## 3. Create another categorical variable named “smoke_gas_exposure” that summarizes “Second Hand Smoke” and “Gas Stove.” The variable should have four categories in total.
```{r}
chs$smoke_gas_exposure <- as.factor(fifelse(chs$smoke == 1, 
                                            fifelse(chs$gasstove == 1, "smoke_gas", "smoke", na=NA) ,
                                            fifelse(chs$gasstove == 1, "gas", "neither", na=NA), na=NA))
summary(chs$smoke_gas_exposure)
```

## 4. Create four summary tables showing the average (or proportion, if binary) and sd of “Forced expiratory volume in 1 second (ml)” (an asthma indicator) by town, sex, obesity level, and “smoke_gas_exposure.”
```{r}
# by town
fev_town <- chs[ , .(
  mean_fev = mean(fev, na.rm = TRUE),
  sd_fev = sd(fev, na.rm = TRUE),
  prop_asthma = sum(asthma==1, na.rm = TRUE)/sum(asthma==1 | asthma==0, na.rm = TRUE),
  sd_asthma = sd(asthma, na.rm = TRUE))
, by = "townname"]
knitr::kable(fev_town)


# by sex
fev_sex <- chs[ , .(
  mean_fev = mean(fev, na.rm = TRUE),
  sd_fev = sd(fev, na.rm = TRUE),
  prop_asthma = sum(asthma==1, na.rm = TRUE)/sum(asthma==1 | asthma==0, na.rm = TRUE),
  sd_asthma = sd(asthma, na.rm = TRUE))
, by = "male"]
knitr::kable(fev_sex)

# by obesity_level
fev_obesity_level <- chs[ , .(
  mean_fev = mean(fev, na.rm = TRUE),
  sd_fev = sd(fev, na.rm = TRUE),
  prop_asthma = sum(asthma==1, na.rm = TRUE)/sum(asthma==1 | asthma==0, na.rm = TRUE),
  sd_asthma = sd(asthma, na.rm = TRUE))
, by = "obesity_level"]
knitr::kable(fev_obesity_level)

# by smoke_gas_exposure
fev_smoke_gas_exposure <- chs[ , .(
  mean_fev = mean(fev, na.rm = TRUE),
  sd_fev = sd(fev, na.rm = TRUE),
  prop_asthma = sum(asthma==1, na.rm = TRUE)/sum(asthma==1 | asthma==0, na.rm = TRUE),
  sd_asthma = sd(asthma, na.rm = TRUE))
, by = "smoke_gas_exposure"]
knitr::kable(fev_smoke_gas_exposure)
```


# Look at the data(EDA)
## The primary questions of interest are:
What is the association between BMI and FEV (forced expiratory volume)?
What is the association between smoke and gas exposure and FEV?
What is the association between PM2.5 exposure and FEV?
Follow the EDA checklist from week 3 and the previous assignment. Be sure to focus on the key variables.
```{r}
# check the dimensions, headers, footers, variable names and variable types
# dimensions
dim(chs)
# headers
head(chs)
# footers
tail(chs)
# variable names and types
str(chs)

# key variables
summary(chs$obesity_level)
summary(chs$bmi)
summary(chs$fev)
summary(chs$pm25_mass)
```
There is nothing strange about the key variables and no missing values.

## Exploratory graphs
```{r}
# histograms to look at the distributions over the whole state
hist(chs$bmi)
hist(chs$fev)
```
### BMI: there is a long end tail, the distribution of BMI is skewed.
### FEV: The distribution of fev almost follows the normal distribution.

```{r}
# bar chart
ggplot(chs, aes(x=townname, y=pm25_mass, fill=townname)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="Bar Chart of pm2.5 by Townname", 
       x="Town Name", 
       y="pm2.5 Value") +
  theme(axis.text.x=element_blank())
```
### Mira Loma has the highest pm25_mass and Lompoc has the lowest.

# Visualization
Create the following figures and interpret them. Be sure to include easily understandable axes, titles, and legends.

## 1. Facet plot showing scatterplots with regression lines of BMI vs FEV by “townname”.
```{r}
chs %>% 
  ggplot(mapping = aes(x = bmi, y = fev)) + 
  geom_point(mapping = aes(color = townname)) + 
  geom_smooth(method = lm, mapping = aes(linetype = townname)) +
  labs(title = "Scatterplot of BMI vs. FEV by Townname",
       x = "BMI",
       y = "FEV") +
  facet_wrap(~ townname)
```
### In each town, there is a positive association between BMI and FEV.

## 2. Stacked histograms of FEV by BMI category and FEV by smoke/gas exposure. Use different color schemes than the ggplot default.
```{r}
# by BMI category
ggplot(chs, aes(x = fev, fill = obesity_level)) +
  geom_histogram(binwidth = 50, position = "stack") +
  labs(title = "Stacked Histogram of FEV by BMI Category",
       x = "FEV",
       y = "Count") +
  scale_fill_brewer(palette = "Set1", name = "BMI Category") 
```
### Underweight population tend to have lower FEV. Overweight and obese population tend to have higher FEV.

```{r}
# by smoke/gas exposure
ggplot(chs, aes(x = fev, fill = smoke_gas_exposure)) +
  geom_histogram(binwidth = 50, position = "stack") +
  labs(title = "Stacked Histogram of FEV by Smoke/gas Exposure",
       x = "FEV",
       y = "Count") +
  scale_fill_brewer(palette = "Set1", name = "Smoke/gas Exposure") 
```
### These four smoke/gas exposure categories share the similar distributions of FEV.

## 3. Barchart of BMI by smoke/gas exposure.
```{r}
ggplot(chs, aes(x=smoke_gas_exposure, y=bmi, fill=smoke_gas_exposure)) +
  geom_bar(stat = "summary", fun = "mean", position="dodge") +
  labs(title="Bar Chart of Mean BMI by Smoke/gas Exposure", 
       x="Smoke/gas Exposure Category", 
       y="Mean BMI") +
  scale_fill_brewer(palette="Dark2")

```
### Overall, the difference of mean BMI is very small between each two smoke/gas exposure categories. 

## 4.Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category.
```{r}
# by BMI Category
chs %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = obesity_level, y = fev, fill = obesity_level)) +
  labs(title = "FEV by BMI Category", x  = "Obesity Level", y = "FEV")

```
### The underweight population has the lowest FEV and the obese population has the highest. The relationship between FEV and BMI categories is positive.

```{r}
# by smoke/gas exposure
chs %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = smoke_gas_exposure, y = fev, fill = smoke_gas_exposure)) +
  labs(title = "FEV by Smoke/gas Exposure Category ", x  = "Smoke/gas Exposure Category", y = "FEV") 
```
### There is no obvious difference of FEV among these four Smoke/gas Exposure categories.

## 5.A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities.
```{r}

pal <- colorFactor( palette = "inferno", domain = chs$pm25_mass)

leaflet(chs) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircleMarkers(lat =~ lat,
             lng =~ lon,
             opacity = 0.5, 
             fillOpacity = 0.5, radius = 5, color =~ pal(pm25_mass),
             popup = ~paste("Community: ", townname, "<br>PM2.5 Mass: ", pm25_mass, " µg/m³"),
  ) %>%
  addLegend(position = "bottomleft", pal = pal, values = chs$pm25_mass,
            title = 'Concentrations of PM2.5 mass among CHS communities', opacity = 1
            )
```
### The map shows that the communities closer to the downtown area have higher PM2.5_mass.

## 6. Choose a visualization to examine whether PM2.5 mass is associated with FEV.
```{r}
# Since pm25_mass and fev are both numeric variables, we choose scatter plot
ggplot(chs, aes(x = pm25_mass, y = fev)) +
  geom_point() +               
  geom_smooth(method = "lm") +
  labs(title = "Scatterplot of PM2.5 Mass vs. FEV",
       x = "PM2.5 Mass (µg/m³)",
       y = "FEV")
```
### We find that pm25_mass only have 12 values, so we can treat it as a categorical variable and make a boxplot.
```{r}
# boxplot
ggplot(chs, aes(x=factor(chs$pm25_mass), y=chs$fev)) +
  geom_boxplot() +
  labs(title="Boxplot of FEV across different PM2.5 Mass levels", 
       x="PM2.5 Mass", 
       y="FEV") +
  theme_minimal()
```
### From the scatter plot, it seems that PM2.5 mass is slightly negatively associated with FEV. But the association is not obvious in the boxplot.













